---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Fasano Franceschini Test

<!-- badges: start -->
<!-- badges: end -->

The fasano.franceschini.test package is an R implementation of the 2-D Kolmogorov-Smirnov (**KS**) two-sample test as defined by Fasano and Franceschini (1987). This is a variant of the 2-D two-sample KS test as originally defined by Peacock (1983).

## Installation

You can install the released version of fasano.franceschini.test from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("fasano.franceschini.test")
```

And the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("nesscoder/fasano.franceschini.test")
```
## Example

#### Underlying Distributions Are Different

```{r}
library(fasano.franceschini.test)

#set seed for reproducible example
set.seed(123)

#create 2-D samples with the different underlying distributions
sample1Data <- data.frame(
  x = rnorm(n = 100, mean = 0, sd = 3),
  y = rnorm(n = 100, mean = 0, sd = 1)
)
sample2Data <- data.frame(
  x = rnorm(n = 100, mean = 0, sd = 1),
  y = rnorm(n = 100, mean = 0, sd = 3)
)

fasano.franceschini.test(sample1Data,sample2Data)
```

#### Underlying Distributions Are The Same

```{r}
#set seed for reproducible example
set.seed(123)

#create 2-D samples with the same underlying distributions
sample1Data <- data.frame(
  x = rnorm(n = 100, mean = 0, sd = 1),
  y = rnorm(n = 100, mean = 0, sd = 1)
)
sample2Data <- data.frame(
  x = rnorm(n = 100, mean = 0, sd = 1),
  y = rnorm(n = 100, mean = 0, sd = 1)
)

fasano.franceschini.test(sample1Data,sample2Data)
```
